This chapter instantiates the methodological framework of Chapter~3 on concrete datasets,
model families, and experimental settings. We first specify the core experimental setup:
datasets, data splitting, model families, preprocessing, Rashomon-set approximation, spatial parameters, and null-model configuration. We then describe
the additional analyses used to assess robustness, validate the pipeline, and interpret the
resulting hotspot structure. Unless stated otherwise, all design choices are fixed across
experiments.

\subsection{Datasets}

Experiments are conducted on three real-world benchmark datasets from different
application domains. All three tasks are binary classification problems and contain a mix
of numerical and categorical input features.

\begin{itemize}
    \item \textbf{COMPAS}: The ProPublica COMPAS recidivism dataset
    \citep{angwin_machine_2016, larson_how_2016}. It contains defendant-level
    information from Broward County, Florida, such as age, prior offence count, sex,
    race, and charge degree. The target indicates whether the individual reoffended
    within two years.

    \item \textbf{German Credit}: The \texttt{credit-g} dataset from OpenML
    \citep{hofmann_statlog_1994}. It contains financial and demographic information about
    credit applicants. The target indicates good versus bad credit risk.

    \item \textbf{Adult}: The Adult income dataset from OpenML \citep{becker_adult_1996}.
    It contains demographic and employment-related information. The target indicates
    whether annual income exceeds \$50{,}000.
\end{itemize}

The datasets differ in size, domain, and feature composition, which makes them useful for
checking whether the proposed analysis behaves similarly across different tabular
classification settings.

\subsection{Data Splitting}
\label{sec:exp-splits}

Each dataset is split into training (60\%), validation (20\%), and test (20\%) sets using
stratified sampling. The three splits have different roles. The training set is used to fit
all candidate models. The validation set is used to rank these candidates by Brier score
and to construct the finite top-$K$ model set $\mathcal{R}_K$. The test set is held out
until the final analysis step and is used only to compute prediction instability and the
spatial statistics.

The full experiment is repeated with 10 different random seeds. This is done to check
whether the results depend strongly on one particular train--validation--test split. Two
splitting modes are used, because the main analyses and the stability analyses require
slightly different comparisons.

\begin{description}
  \item[Independent splits.] For the main analyses, each random seed creates a new
  stratified train--validation--test split. Thus, the training, validation, and test sets all
  differ across runs. This mode reflects the variability of the full experimental pipeline:
  model fitting, validation-based model selection, test-set prediction, and spatial
  analysis.

  \item[Fixed-test-set splits.] For the cross-seed stability analysis, the test set is drawn
  once and kept fixed across all runs. Only the training and validation parts are re-split.
  This makes it possible to compare hotspot membership for the same test observations
  across runs. Without a fixed test set, pointwise comparisons such as HH selection
  frequency or Jaccard overlap would not be meaningful, because the test observations
  would change from run to run.
\end{description}

In both splitting modes, the test set is never used for training or for selecting the
top-$K$ model set. It is used only after model selection, when the selected models are
applied to the test observations and the multiplicity and spatial analyses are computed.

\subsection{Model Families and Training}

All models are implemented in Python using \texttt{scikit-learn}
\citep{pedregosa2011scikit}. The experiments use five model families in order to
cover different types of classifiers: a linear model, an instance-based method, two
tree-based ensemble methods, and a neural-network model.

\begin{itemize}
    \item \textbf{Logistic Regression (LogReg):} a linear probabilistic classifier with
    elastic-net regularization.

    \item \textbf{$k$-Nearest Neighbors (kNN):} an instance-based classifier whose
    predictions depend on the labels of nearby training observations.

    \item \textbf{Random Forest (RF):} an ensemble of decision trees trained with
    bootstrap aggregation.

    \item \textbf{Gradient Boosting (GBM):} an additive ensemble of shallow decision
    trees trained sequentially.

    \item \textbf{Multilayer Perceptron (MLP):} a feed-forward neural network trained
    with the Adam optimizer and early stopping.
\end{itemize}

For each model family and run, 50 hyperparameter configurations are sampled from
predefined finite search spaces. The full search spaces are reported in
Appendix~\ref{app:hp-search-space}. Configurations are sampled using deterministic seeds
derived from the run, so that the experiment is reproducible. If duplicate
configurations occur for families with smaller discrete search spaces, configurations are
resampled until 50 unique candidates are obtained.

Each candidate model is trained on the training set only. The validation set is used to
rank the candidates by Brier score and to construct the finite top-$K$ model set described
in Section~\ref{sec:meth-rashomon}. The test set is not used for model selection; it is
used only for computing prediction instability and the subsequent spatial analyses. No
cross-validation is used inside an run; instead, the whole pipeline is repeated across
10 outer train--validation--test splits.

\subsection{Preprocessing}
\label{sec:exp-preproc}

Preprocessing is applied uniformly across all model families via a column transformer fitted
on the training set only.

\begin{itemize}
  \item \textbf{Numeric features:} missing values are imputed with the training-set median, then standardized to zero mean and unit variance.
  \item \textbf{Categorical features:} missing values are imputed with the training-set mode, then one-hot encoded. Unknown categories at validation or test time are mapped to a zero vector.
  \item \textbf{Remaining columns:} any column not identified as numeric or categorical is dropped.
\end{itemize}

The fitted transformer is applied identically to validation and test data without refitting,
thereby avoiding data leakage. The same transformed feature space is used both for model
training and for constructing the kNN graph (Section~\ref{sec:exp-knn-graph}), so that
spatial neighborhoods reflect the representation on which the models operate.
\subsection{Rashomon Set Approximation}
\label{sec:exp-rashomon}

In the experiments, the Rashomon set is approximated by selecting the \textbf{top-$K$}
models ranked by validation Brier score, with $K = 25$ unless otherwise stated. Two
selection modes are used:

\begin{itemize}
    \item \textbf{Global:} the 25 models with the lowest validation Brier score from the full candidate pool, with all model families combined.
    \item \textbf{Per-family ($K$-each):} the top 25 models are selected separately within each model family. This mode is used for family-wise comparisons and for within-family driver analyses, because it retains sufficient variation within each family.
\end{itemize}

Unless otherwise stated, the global selection mode is used. Sensitivity to the Rashomon-set
size is examined by varying $K \in \{5, 10, 15, 20, 25, 30, \ldots\}$ in separate analyses.
Results are reported for the global Rashomon approximation and, where relevant, for the
per-family $K$-each construction (Sections~\ref{sec:results-global-vs-family}
and~\ref{sec:results-hp}).

\subsection{Prediction Matrix and Multiplicity Metrics}

For each run, all models in the selected Rashomon approximation are applied to the held-out
test set to obtain predicted probabilities. This yields a prediction matrix
$P \in \mathbb{R}^{K \times N}$, where $K$ is the number of selected models and $N$ is
the number of test observations.

Observation-wise predictive variance $v_i$ is computed as defined in Chapter~3. In
addition, we report several aggregate multiplicity summaries: mean variance, ambiguity
(mean over observations of $\max_m \hat{p}_{mi} - \min_m \hat{p}_{mi}$), disagreement
rate (fraction of test points for which model predictions differ by more than
$\delta = 0.05$), and discrepancy (maximum over model pairs of the mean absolute
prediction difference).

\subsection{Spatial Analysis Parameters}
\label{sec:exp-knn-graph}

The kNN graph (Section~\ref{sec:meth-graph}) is constructed from the preprocessed
test-set feature vectors. Rather than using a single fixed neighborhood size across all
datasets, we select dataset-specific values of $k$ so that the graph is sufficiently connected
while still preserving meaningful local structure. In particular, we use $k = 30$ for
COMPAS and German Credit, and a larger value of $k = 60$ for Adult when smaller
values lead to disconnected graph components.
Following Section~\ref{sec:method-spatial}, the initial directed $k$NN adjacency is symmetrized before row-standardization. The reported values of $k$ refer to the initial directed nearest-neighbor query; after symmetrization, individual nodes may have more than $k$ neighbors.

Distances are Euclidean and the adjacency matrix is row-standardized. Global Moran's $I$
significance is assessed with 999 permutations. Local Moran (LISA) also uses 999
permutations per observation, and the Benjamini--Hochberg procedure is applied at
$\alpha = 0.05$ to control the false discovery rate when labeling HH and LL clusters.
Connected components of HH observations are extracted from the restricted kNN graph;
where stated, a minimum component size is used to suppress very small fragments.

\subsection{Null Experiment Configuration}
\label{sec:exp-null}

The permutation null follows the procedure defined in Section~\ref{sec:meth-null}. In
the experiments, we use $R = 100$ null permutations per run. For each permutation, predictive variance is recomputed from the permuted prediction
matrix, and global Moran's $I$ is recalculated.

A run is treated as significant when the empirical permutation $p$-value is below 0.05.
Because $R=100$, the smallest attainable empirical $p$-value is $1/(R+1)=1/101
\approx 0.0099$. Reported values of ``$p<0.01$'' should therefore be interpreted at this
resolution limit.

\subsection{Runs and Reproducibility}

For each dataset, 10 runs are performed with random seeds 0--9. Each run
corresponds to a complete train/validation/test split and a full training, selection, and
analysis pipeline. Results are aggregated across runs, typically by reporting means and
standard deviations of quantities such as mean variance, Moran's $I$, HH counts, and the
fraction of runs with significant Moran's $I$. All random seeds are fixed for reproducibility.


\subsection{Robustness and Validation Analyses}

The following additional analyses are used to assess whether the observed spatial structure
is robust, methodologically stable, and recoverable under controlled conditions.

\paragraph{Calibration robustness.}
\label{sec:exp-calibration}
To test whether spatial multiplicity is driven by probability miscalibration, we apply Platt
scaling and isotonic regression to the predictions of each model in the Rashomon set. For each model, a logistic
regression is fitted on validation-set predicted probabilities and validation labels, and the
resulting mapping is applied to that model's test-set predictions. The full multiplicity and
spatial pipeline is then rerun on both uncalibrated and calibrated test predictions. We
record changes in mean variance, Moran's $I$, HH/LL masks, connected components, and
the Jaccard similarity between hotspot masks before and after calibration. Synthetic
datasets are excluded from this analysis because they follow a separate data-generating
design.

\paragraph{Alternative graph constructions.}
To verify that the spatial results are not artifacts of Euclidean distance on the full one-hot
encoded feature space, we compare the baseline graph with two alternatives:
(i) kNN in PCA-reduced space (retaining 5 components), and
(ii) kNN using cosine distance on $\ell_2$-normalized features. All other pipeline
parameters remain fixed, and Moran's $I$ and HH counts are compared across graph
definitions.

\paragraph{Synthetic validation.}
Two synthetic data designs are used to validate whether LISA-based HH regions recover
known ambiguous regions:
\begin{itemize}
    \item \textbf{Single-island:} a mixture of two well-separated Gaussian blobs (stable region) and one ambiguous island, with points drawn uniformly in a disk and weak XOR-like conditional probability $P(y=1\mid x) = 0.5 \pm \delta$.
    \item \textbf{Three-islands + outliers:} two stable blobs, three ambiguous islands, and a small fraction of isolated outliers, with known region labels.
\end{itemize}
The same training pipeline (60/20/20 split, same model families, top-$K$ by validation
Brier score) is applied. Recovery of the ground-truth ambiguous regions by LISA HH
classification is evaluated using precision, recall, and Jaccard overlap. The null
experiment from Section~\ref{sec:exp-null} is used to verify that observed Moran's $I$ is
significant.

\subsection{Interpretation and Attribution Analyses}

The following analyses are used to identify which modeling choices drive predictive multiplicity and to characterize hotspot regions in terms of original features. The attribution analyses are descriptive rather than causal: they quantify how disagreement varies with model family and hyperparameter choices within the sampled candidate pool.

\paragraph{Family importance.}

We first decompose predictive variance into a between-family component and a
within-family component. For each dataset and random seed, we construct one
family-specific near-optimal model set per model family by selecting the top 25 models
within that family. Let \(\mathcal{G}\) denote the set of model families, and let
\(\mathcal{R}^{(g)}_K\) denote the top-\(K\) model set for family \(g \in \mathcal{G}\).
For observation \(i\), let \(\hat{p}_{gmi}\) be the predicted probability from model
\(m \in \mathcal{R}^{(g)}_K\). The mean prediction of family \(g\) is

\[
\bar{p}_{gi}
=
\frac{1}{K}\sum_{m \in \mathcal{R}^{(g)}_K} \hat{p}_{gmi},
\]

and the overall mean prediction across all family-specific model sets is

\[
\bar{p}_{i}
=
\frac{1}{|\mathcal{G}|}\sum_{g \in \mathcal{G}} \bar{p}_{gi}.
\]

For a set of observations \(S\), for example all test observations, HH observations, or
non-HH observations, the between-family component is computed as

\[
\mathrm{Var}_{\mathrm{between}}(S)
=
\frac{1}{|S|}\sum_{i \in S}
\frac{1}{|\mathcal{G}|}
\sum_{g \in \mathcal{G}}
\left(\bar{p}_{gi} - \bar{p}_{i}\right)^2 .
\]

This term is large when different model families have different average predictions for
the same observations. The total predictive variance across all selected models is

\[
\mathrm{Var}_{\mathrm{total}}(S)
=
\frac{1}{|S|}\sum_{i \in S}
\frac{1}{|\mathcal{G}|K}
\sum_{g \in \mathcal{G}}
\sum_{m \in \mathcal{R}^{(g)}_K}
\left(\hat{p}_{gmi} - \bar{p}_{i}\right)^2 .
\]

Family importance is then defined as

\[
\mathrm{FI}(S)
=
\frac{\mathrm{Var}_{\mathrm{between}}(S)}
{\mathrm{Var}_{\mathrm{total}}(S)}.
\]

A value close to one means that most of the prediction variance is explained by
differences between model families. A lower value means that a larger share of the
variation remains within families, for example due to different hyperparameter settings
inside the same model family. Because each family contributes the same number of models
in this analysis, the decomposition gives equal weight to each model family. We compute
\(\mathrm{FI}(S)\) for all test observations and separately for HH and non-HH observations
to check whether family choice is more important inside high-multiplicity regions.

\paragraph{Within-family hyperparameter importance.}

After separating the between-family component, we analyze which hyperparameters explain
the remaining disagreement within each model family. Let \(g \in \mathcal{G}\) denote a
model family, and let \(\mathcal{R}^{(g)}_K\) be the top-\(K\) model set for that family.
For observation \(i\), let \(\bar{p}^{(g)}_i\) denote the mean prediction across models in
family \(g\):
\[
\bar{p}^{(g)}_i
=
\frac{1}{K}
\sum_{m \in \mathcal{R}^{(g)}_K}
\hat{p}^{(g)}_{mi}.
\]

For each model \(m \in \mathcal{R}^{(g)}_K\), we compute its model-level disagreement
contribution as
\[
V^{(g)}_m =
\frac{1}{N_{\mathrm{test}}}
\sum_{i=1}^{N_{\mathrm{test}}}
\left(
\hat{p}^{(g)}_{mi} - \bar{p}^{(g)}_i
\right)^2 .
\]
Thus, \(V^{(g)}_m\) measures how strongly model \(m\)'s predictions deviate from the
average prediction of its own family.

For each hyperparameter \(h\) within family \(g\), models are grouped according to the
observed value or range of that hyperparameter. Categorical hyperparameters and numeric
hyperparameters with at most four distinct values are grouped by exact value. Numeric
hyperparameters with more than four distinct values are discretized into three quantile
bins. If the resulting groups have insufficient support, or fewer than two valid groups
remain, the hyperparameter is excluded for that seed.

A one-way variance decomposition of \(V^{(g)}_m\) across the groups of hyperparameter
\(h\) yields a between-group component
\(\mathrm{Var}_{\mathrm{between}}^{(g)}(h)\) and a total component
\(\mathrm{Var}_{\mathrm{total}}^{(g)}\). Hyperparameter importance is defined as
\[
\rho^{(g)}(h)
=
\frac{\mathrm{Var}_{\mathrm{between}}^{(g)}(h)}
{\mathrm{Var}_{\mathrm{total}}^{(g)}}.
\]
The procedure is repeated for each family and random seed. Results are then aggregated by
mean importance, standard deviation, and the number of valid seeds.

\paragraph{Hotspot-specific hyperparameter shifts.}
To assess whether hyperparameter effects differ inside multiplicity hotspots, the same \(V_m\)-based importance calculation is repeated on HH observations. For each hyperparameter, we compare its hotspot-specific importance to its global importance:
\[
\Delta \rho(h)
=
\rho_{\mathrm{HH}}(h) - \rho_{\mathrm{all}}(h).
\]
Positive values indicate that the hyperparameter explains a larger share of within-family disagreement inside HH regions, while negative values indicate lower importance inside hotspots. Because HH subsets can be small, especially on German Credit, these estimates are interpreted descriptively and are reported with seed coverage where relevant.


\paragraph{Interpretable rule extraction.}
To characterize high-multiplicity regions in terms of original features, interpretable rules
are extracted using shallow decision trees. The primary target label is defined as
\textbf{HH}, corresponding to LISA High--High hotspot observations.

We apply the rule extraction in two variants. First, for each run, a global rule model is
fitted to predict membership in the full HH set. A decision tree with maximum depth 3 is
trained on the original feature space, and positive leaf nodes are converted into
conjunctive rules. These global rules describe broader subgroups that are enriched for
HH observations, but they may mix several distinct hotspot regions.

Second, we perform a component-level rule analysis. Connected components of the HH
subgraph are treated separately, and for each retained component a shallow decision tree
is fitted with membership in that component as the positive class and all other test
observations as the negative class. This variant is motivated by the stability analysis,
which shows that hotspots are better interpreted as localized regions than as
pointwise-stable labels. Component-level rules therefore aim to describe compact
high-multiplicity regions rather than the full HH set at once.

Rules are evaluated using support (coverage), purity, recall, and lift. For global HH rules,
purity is defined as $P(\text{HH} \mid \text{rule region})$, and lift is measured relative to
the HH base rate. For component-level rules, purity is defined as
$P(\text{component} \mid \text{rule region})$, and lift is measured relative to the base
rate of the corresponding component. Stability is assessed across random seeds by
aggregating rule statistics and analyzing feature frequency. The main rule analysis is
reported for COMPAS, where the hotspot structure is clearest and the retained
components are sufficiently large for meaningful rule extraction.
